---
title: "P004MendelsFirstLaw"
author: "McKayla Ford"
date: "2026-08-11"
output: html_document
---

I don't love this problem, because they're asking for a probability tree
solution, when hardy-weinburg solved this problem more elegantly in a population
genetics context (but offers less precision for the exact numbers desired).

Namely, in the example:
```{r}
K <- 22
M <- 26
N <- 26
```

where k m and n are homozygote dominate, heterozygote, and homozygote recessive
respectively, the expected output for number individuals with a dominant phenotype
is, according to Rosalind, 0.7833.
However, the Hardy-Weinburg calculation would be as follows:
```{r}
tot_pop <- K + M + N
p <- (K+0.5*M)/tot_pop
q <- (N+0.5*M)/tot_pop
perc_dom <- p^2 + 2*p*q
perc_rec <- q^2
```

```{r}
perc_dom
```
Unfortunately that's not an acceptable result - for rosalind here, it is desirable
to model that each parent is being subtracted from the population, even though on
a population genetics scale, that contribution is neglible. Therefore...

There is probability K/tot_pop, M/tot_pop, N/tot_pop of drawing each parent for first parent.
There is then probability K-1/tot_pop-1, M/tot_pop-1, N/tot_pop-1 for dom, same but shift type for each OG parent.
If any parent is K, baby will have a dom allele.
If both parents are N, baby will not have dom allele.
If both parents are M, 3/4 chance dom.
If 1 parent M and 1 parent R, 1/2 chance dom.

Probabilty tree version...
```{r}
c_dom <- K/tot_pop # any dom will have a dom gene
c_hetdom <- M/tot_pop * (K)/(tot_pop-1)
c_resdom <- N/tot_pop * (K)/(tot_pop-1)

c_hethet <- M/tot_pop * (M-1)/(tot_pop-1) * 3/4 #only 3/4 of hethet will have dom

c_hetres <- M/tot_pop * (N)/(tot_pop-1) * 1/2 #only 1/2 of het + res will have dom
c_reshet <- N/tot_pop * (M)/(tot_pop-1) * 1/2
#c resres unnessary, no resxres will have dom gene

chance_has_dom <- c_dom + c_hetdom + c_resdom + c_hethet + c_hetres + c_reshet
chance_has_dom
```
Yeah, there's the answer rosalind wants. Notably as pop size increases difference
to hardy weinburg reduces.

I'm not using fancier packages, just math/logic, which would be the same in 
python, so I'm not going to implement a python version for this one, though I 
will review the python solutions for my education.



